Constraint Grammar as a SAT problem

### 2015-06-01 paper/nodalida2015.tex
stuff

We represent Constraint Grammar (CG) as a Boolean satisfiability 
(SAT) problem.
Encoding CG in logic brings some new features to the grammars.
The rules are interpreted in a more declarative way, which makes it
possible to abstract away from details such as cautious context and
ordering.
The encoding in logic captures richer dependencies between the tags:
a rule is allowed to take action also concerning the words in its
context. This makes the number of the rules potentially smaller. 
Ordering can be preserved or discarded;
in the latter case, we solve eventual rule conflicts by finding a
solution that discards the least number of rule applications.
We test our implementation by parsing texts in the order of
10,000s--100,000s words, using grammars with hundreds of rules.
\end{abstract}

@@

Applying logic to reductionist grammars has been explored earlier by \cite{lager98,lager_nivre01}, but it was never adopted for use.
% ; logic programming was too slow to be used for tagging or parsing. 
Since those works, SAT solving techniques have improved significantly \cite{marques_silva2010}, and they are used in domains such as microprocessor design and computational 
biology---these problems easily match or exceed CG in complexity. 
Due to these advances, we were able to revisit the idea and develop it
further.

@@

With these constraints, we get two solutions. The interaction of the rules regarding \emph{la}  disambiguates \emph{casa} for free, and ordering of the rules doesn't matter.

@@

Traditional CG relies on that applying the rule set iteratively:
some rules fire during the first iteration, either because their
conditions don't require cautious context, or because some words are
unambiguous to start with. This makes some more words unambiguous, and
new rules can fire during the second iteration.

@@

As hinted by the previous property, SAT solver doesn't need a fixed
order of the rules.
Applying a rule to a sentence produces a number of clauses,
and those clauses are fed into the SAT solver.
However, in the unordered scheme, some information is lost: the
following rule sets would be treated identically, whereas in the
traditional CG, only the first would be considered as a bad order.

@@

In case of a conflict, SAT solver makes it possible to discard only
minimal amount of rule applications. Continuing with the example,
it is not clear which instances would be discarded, but if the rules
were part of a larger rule set, and in the context the REMOVE rule was
the right one to choose, it is likely that the interaction between the desired
rules would make a large set of clauses that fit together, and the
SELECT rule would not fit in, hence it would be discarded.

@@

%With the added interaction between rules, catch-all rules might even be less 
%necessary. However, this is just speculation, we have not tested this effect.

%We can apply this for the whole grammar, or for one section at a time.

%The design of unordered rules can lead to new ways of writing CGs,
%potentially with less mental effort to the writer, but also less transparent.

%By testing against gold standard, we see that the
%unordered scheme not optimal for applying ready-made
%CGs, where the order is a crucial detail.

We can also emulate order with SAT-CG. To do that, we enter clauses
produced by each rule one by one, and assume the solver state reached
so far is correct. If a new clause introduces a conflict with
previous clauses, we discard it and move on to the next rule.
By testing against gold standard, we see that this scheme works better
with ready-made CGs, which are written with ordering in mind.
It also runs slightly faster than the unordered version.
%the performance, because the maximisation problems are smaller. 
%We have also experimented with sections, and 
\\

%  that it is possible to write a small grammar, 
% which, when run with the unordered scheme, disambiguates more than the
% same grammar run with VISL CG-3. 
% However, when tested with grammars that are written for the original
% CG, SAT-CG loses with both strategies, but emulating order fares better.

% Finite-state implementations of CG \cite{koskenniemi92} also have logically unordered rules. 
% \todo{Inari: check if Tapanainen 1996 discuss ordering}

\noindent These three features influence the way that rules are written. 
We predict that less rules are needed; whether this holds in the order
of thousands of rules remains to be tested. On one hand, getting rid
of ordering and cautious context could ease the task of the grammar
writer, since it removes the burden of estimating the best sequence of
rules and whether to make the them cautious. On the other
hand, lack of order can make the rules less transparent, and might not scale up for larger grammars.

@@

We took a manually tagged
corpus\footnote{https://svn.code.sf.net/p/apertium/svn/branches/apertium-swpost/apertium-en-es/es-tagger-data/es.tagged}
containing approximately 22,000 words of Spanish news text, 
and a small constraint grammar\footnote{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}, produced independently of the authors.
% We ignored substitute rules, but kept subreadings and unification, even though our implementation doesn't handle them.
We kept only select and remove rules, after which the grammar had 261 rules.
With this setup, we produced an ambiguous version of the tagged
corpus, and ran both SAT-CG and VISL CG-3 on it.
Treating the original corpus as the gold standard, the disambiguation
by  VISL CG-3 achieves F-score of 82.6 \%, ordered SAT-CG 81.5 \%  and
unordered SAT-CG 79.2 \%. 
We did not test with other languages or text genres due to the lack of
available gold standard.
%This result suggests that that when running grammars that are written
%with the traditional CG in mind, SAT-CG loses with both ordering
%strategies, but emulating order fares better.

We also wanted to test the hypothesis that SAT-CG needs different
kinds of rules to perform well. We took the base grammar and started
handcrafting rules; with our best performing and most concise
grammar\footnote{https://github.com/inariksit/cgsat/blob/master/data/spa\_smallset.rlx}
of just 19 rules, unordered SAT-CG achieves 85.3 \% and VISL CG-3 85.9
\%. This experiment is naturally very small and might be explained by
overfitting or mere chance, but it seems to indicate that rules that
work well with SAT-CG, are also good for traditional CG.
% It was possible to attain a rule set for which the unordered SAT-CG
% works better (\textless0.5 \%), but it required trying
% subsequences of the 19-rule set in a brute force manner---this hardly
% presents any real life use case.

@@

The worst-case complexity of SAT is exponential, whereas the standard
implementations of CG are polynomial, but with advances in SAT solving
techniques, the performance in the average case in practice is much more feasible than in the previous works done in 90s--00s.
We used the open-source SAT solver MiniSat \cite{een04sat}.

We tested the performance by parsing Don Quijote (384,155 words) with
the same Spanish grammars as in the previous experiment. 
Table~\ref{table:time} shows execution times compared to VISL CG-3;
SAT-CG\textsubscript{u} is the unordered scheme and
SAT-CG\textsubscript{o} is the ordered.
% For both systems, the number of rules in the grammar affects the performance more than the word count.
From the SAT solving side, maximisation is the most costly operation. 
Emulating order is slightly faster, likely because the maximisation problems are smaller.
In any case, SAT does not seem to be the bottleneck: with 261 rules,
the maximisation function was called 147,253 times, and with 19 rules,
132,255 times, but 
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
the differences in the execution times are much larger, which suggests
that there are other reasons for the worse performance. 
This is to be expected, SAT-CG is currently just a naive
proof-of-concept implementation with no optimisations.

@@

The sequential application of traditional CG rules is good for
performance and transparency. When a rule takes action, the analyses
are removed from the input, and the next rules get the modified sentence
as their input. 
% At the execution of rule, there is no way to go back to earlier rules and undo them. 
As a downside, there is no way to know which part comes directly from the
raw input and which part from applying previous rules.

@@

The first rule selects the verb reading always, removing all other readings,
and the second rule removes the verb reading conditionally.
If the rules are introduced in a different order, there is no
conflict: the REMOVE rule would not remove verb readings from all
possible verb analyses, so there is a possibility for the SELECT rule
to fire.

@@

As future work, we intend to test for conflicts
without using a corpus.
Let us illustrate the idea with the same two rules,
\texttt{SELECT v} and \texttt{REMOVE v IF (-1 det)}
in both orders.
Assume we have the tag set \texttt{\{det, n, v, sg, pl\}}, 
and we want to find if there exists an input such that
 both rules, applied in the given order, remove something from the input.
% This is another constraint solving problem, hence we can model it in SAT.
There are no inputs that satisfy the requirement with first order,
but several that work with the second, for instance

\begin{itemize}
\item []
\begin{verbatim}
"<w1>"
        det
        v
"<w2>"
        n
        v
\end{verbatim}
\end{itemize}
Thus we can say that the first rule set is sane, but the second rule
set is conflicting. 
Implementing and testing this on a larger scale is left for future work.

@@

SAT-solvers are nowadays powerful enough to be used for dealing with
Constraint Grammar. A logic-based approach to CG has possible
advantages over more traditional approaches; a SAT solver may
disambiguate more words, and may do so more precisely, capturing
more dependencies between tags.
We experimented with both ordered and unordered rules, and found the ordered
scheme to work better with previously written grammars.
The evaluation shows that SAT-CG isn't a match for
As future work, we 
As a complementary approach, we intend to concentrate on grammar analysis

>>>>

We represent Constraint Grammar (CG) as a Boolean satisfiability 
(SAT) problem.
Encoding CG in logic brings some new features to the grammars.
The rules are interpreted in a more declarative way, which makes it
possible to abstract away from details such as cautious context and
ordering.
%In addition to the target words, 
A rule is allowed to affect its context words, which makes the number
of the rules in a grammar potentially smaller. 
Ordering can be preserved or discarded;
in the latter case, we solve eventual rule conflicts by finding a
solution that discards the least number of rule applications.
We test our implementation by parsing texts in the order of
10,000s--100,000s words, using grammars with hundreds of rules.
\end{abstract}

@@

Applying logic to reductionist grammars has been explored earlier by \cite{lager98,lager_nivre01}, but it was never adopted for use.
% ; logic programming was too slow to be used for tagging or parsing. 
Since those works, SAT solving techniques have improved significantly \cite{marques_silva2010}, and they are used in domains such as microprocessor design and computational 
biology---these problems easily match or exceed CG in complexity. 
Thanks to these advances, we were able to revisit the idea and develop it
further.

@@

With these constraints, we get two solutions. The interaction of the
rules regarding \emph{la}  disambiguates the part of speech of
\emph{casa} for free, and the order of the rules does not matter.

@@

Traditional CG relies on that applying the rule set iteratively:
some rules fire during the first iteration, either because their
conditions do not require cautious context, or because some words are
unambiguous to start with. This makes some more words unambiguous, and
new rules can fire during the second iteration.

@@

As hinted by the previous property, SAT solver does not need a fixed
order of the rules.
Applying a rule to a sentence produces a number of clauses,
and those clauses are fed into the SAT solver.
However, in the unordered scheme, some information is lost: the
following rule sets would be treated identically, whereas in the
traditional CG, only the first would be considered as a bad order.

@@

In case of a conflict, the SAT solver makes it possible to discard only
minimal amount of rule applications. Continuing with the example,
it is not clear which instances would be discarded, but if the rules
were part of a larger rule set, and in the context the REMOVE rule was
the right one to choose, it is likely that the interaction between the desired
rules would make a large set of clauses that fit together, and the
SELECT rule would not fit in, hence it would be discarded.

@@

We can also emulate order with SAT-CG. To do that, we enter clauses
produced by each rule one by one, and assume the solver state reached
so far is correct. If a new clause introduces a conflict with
previous clauses, we discard it and move on to the next rule.
By testing against gold standard, we see that this scheme works better
with ready-made CGs, which are written with ordering in mind.
It also runs slightly faster than the unordered version.
\\

\noindent These three features influence the way rules are written. 
We predict that less rules are needed; whether this holds in the order
of thousands of rules remains to be tested. On the one hand, getting rid
of ordering and cautious context could ease the task of the grammar
writer, since it removes the burden of estimating the best sequence of
rules and whether to make them cautious. On the other
hand, lack of order can make the rules less transparent, and might not scale up for larger grammars.

@@

We took a manually tagged
corpus\footnote{https://svn.code.sf.net/p/apertium/svn/branches/apertium-swpost/apertium-en-es/es-tagger-data/es.tagged}
containing approximately 22,000 words of Spanish news text, 
and a small constraint grammar\footnote{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}, produced independently of the authors.
% We ignored substitute rules, but kept subreadings and unification, even though our implementation doesn't handle them.
We kept only SELECT and REMOVE rules, which left us 261 rules.
With this setup, we produced an ambiguous version of the tagged
corpus, and ran both SAT-CG and VISL CG-3 on it.
Treating the original corpus as the gold standard, the disambiguation
by  VISL CG-3 achieves F-score of 82.6 \%, ordered SAT-CG 81.5 \%  and
unordered SAT-CG 79.2 \%. 
We did not test with other languages or text genres due to the lack of
available gold standard.
%This result suggests that that when running grammars that are written
%with the traditional CG in mind, SAT-CG loses with both ordering
%strategies, but emulating order fares better.

We also wanted to test the hypothesis that SAT-CG works better with
small amount of rules. We took the base grammar and started
handcrafting rules; with our best performing and most concise
grammar\footnote{https://github.com/inariksit/cgsat/blob/master/data/spa\_smallset.rlx}
of only 19 rules, unordered SAT-CG achieves 85.3 \% and VISL CG-3 85.9
\%. This experiment is naturally very small and might be explained by
overfitting or mere chance, but it seems to indicate that rules that
work well with SAT-CG, are also good for traditional CG.
% It was possible to attain a rule set for which the unordered SAT-CG
% works better (\textless0.5 \%), but it required trying
% subsequences of the 19-rule set in a brute force manner---this hardly
% presents any real life use case.

@@

The worst-case complexity of SAT is exponential, whereas the standard
implementations of CG are polynomial, but with advances in SAT solving
techniques, the performance in the average case in practice is more feasible than in the previous works done in 90s--00s.
We used the open-source SAT solver MiniSat \cite{een04sat}.

We tested the performance by parsing Don Quijote (384,155 words) with
the same Spanish grammars as in the previous experiment. 
Table~\ref{table:time} shows execution times compared to VISL CG-3;
SAT-CG\textsubscript{u} is the unordered scheme and
SAT-CG\textsubscript{o} is the ordered.
% For both systems, the number of rules in the grammar affects the performance more than the word count.
From the SAT solving side, maximisation is the most costly operation. 
Emulating order is slightly faster, likely because the maximisation problems are smaller.
In any case, SAT does not seem to be the bottleneck: with 261 rules,
the maximisation function was called 147,253 times, and with 19 rules,
132,255 times, but 
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
the differences in the execution times are much larger, which suggests
that there are other reasons for the worse performance. 
This is to be expected, as SAT-CG is currently just a naive
proof-of-concept implementation with no optimisations.

@@

The sequential application of traditional CG rules is good for
performance and transparency. When a rule takes action, the analyses
are removed from the sentence, and the next rules get the modified
sentence as input. 
% At the execution of rule, there is no way to go back to earlier rules and undo them. 
As a downside, there is no way to know which part comes directly from the
raw input and which part from applying previous rules.

@@

The first rule selects the verb reading everywhere and removes
all other readings, leaving no chance for the second rule to take action.
If the rules are introduced in a different order, there is no
conflict: the REMOVE rule would not remove verb readings from all
possible verb analyses, so there is a possibility for the SELECT rule
to fire.

@@

As future work, we intend to test for conflicts
without using a corpus.
Let us illustrate the idea with the same two rules,
\texttt{SELECT v} and \texttt{REMOVE v IF (-1 det)}
in both orders.
Assume we have the tag set \texttt{\{det, n, v, sg, pl\}}, 
and we want to find if there exists an input such that
 both rules, applied in the given order, remove something from the input.
% This is another constraint solving problem, hence we can model it in SAT.
There are no inputs that satisfy the requirement with the first order,
but several that work with the second, such as the following:

\begin{itemize}
\item []
\begin{verbatim}
"<w1>"
        det
        v
"<w2>"
        n
        v
\end{verbatim}
\end{itemize}
Thus we can say that the first rule order is conflicting, but the second
one is not. 
Implementing and testing this on a larger scale is left for future work.

@@

SAT-solvers are nowadays powerful enough to be used for dealing with
Constraint Grammar. A logic-based approach to CG has possible
advantages over more traditional approaches; a SAT solver may
disambiguate more words, and may do so more precisely, capturing
more dependencies between tags.
We experimented with both ordered and unordered rules, and found the 
ordered scheme to work better with previously written grammars.
For future direction, we intend to concentrate on grammar analysis,
especially finding conflicts in grammars.
